---
title: "Home Credit Default Risk: Exploratory Data Analysis"

format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    embed-resources: true

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: true

needed <- c("tidyverse", "janitor", "skimr", "naniar", "scales")
to_install <- needed[!needed %in% installed.packages()[, "Package"]]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(scales)

theme_set(theme_minimal())
options(scipen = 999)
```
## 1. Project Context

This notebook presents exploratory data analysis for the Home Credit Default Risk project. The purpose of this analysis is to better understand the borrower data, identify data quality issues, and examine patterns related to default risk. These findings help guide later preprocessing, feature engineering, and modeling decisions.

## 2. Load Data

```{r}
#| label: load-data
#| echo: true

# Update this path if your file is stored somewhere else
data_path <- "application_train.csv"

app_train <- read_csv(data_path) |> clean_names()
```

```{r}
#| label: first-look
#| echo: true

head(app_train)
```
The dataset was loaded successfully and cleaned so that variable names are in a consistent format. This makes later analysis easier and reduces errors caused by formatting differences in column names.

## 3. Data Overview

```{r}
#| label: data-dimensions
#| echo: true

dim(app_train)
```

```{r}
#| label: column-names
#| echo: true

names(app_train)[1:20]
```

```{r}
#| label: data-summary
#| echo: true

skim(app_train)
```
This overview gives a general understanding of the size and structure of the dataset. It also provides an early look at missing values, variable types, and possible quality issues that will affect downstream analysis.

## 4. Target Distribution

```{r}
#| label: target-counts
#| echo: true

app_train |>
  count(target) |>
  mutate(percent = n / sum(n))
```

```{r}
#| label: target-plot
#| echo: true

app_train |>
  count(target) |>
  ggplot(aes(x = factor(target), y = n)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Default Target",
    x = "Target (0 = Non-default, 1 = Default)",
    y = "Number of Applicants"
  )
```
The target variable is clearly imbalanced, with many more non-default cases than default cases. This is important because class imbalance can affect both model training and evaluation. It suggests that model performance should not be judged only by overall accuracy.

## 5. Missing Data Analysis

```{r}
#| label: missing-summary
#| echo: true

missing_summary <- app_train |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "missing_pct") |>
  mutate(missing_pct = missing_pct * 100) |>
  arrange(desc(missing_pct))

head(missing_summary, 20)
```

```{r}
#| label: missing-plot
#| echo: true
#| fig-height: 7
#| fig-width: 10

missing_summary |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(variable, missing_pct), y = missing_pct)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 20 Variables by Missing Percentage",
    x = "Variable",
    y = "Missing Percentage"
  )
```

```{r}
#| label: missing-visual
#| echo: true
#| fig-height: 6
#| fig-width: 10

vis_miss(app_train |> select(1:20))
```
Several variables contain substantial missingness. This indicates that missing-value handling will be a major part of the preprocessing workflow. Features with extremely high missingness may require careful imputation or may need to be excluded, depending on their usefulness.

## 6. Basic Feature Cleanup for Exploration
Some variables in this dataset are easier to interpret after simple transformations. In particular, age and employment duration are stored as negative day counts, and the special employment value 365243 appears to represent an anomalous or placeholder value. 

```{r}
#| label: basic-cleanup
#| echo: true

eda_df <- app_train |>
  mutate(
    years_birth = round(abs(days_birth) / 365, 1),
    days_employed = if_else(days_employed == 365243, NA_real_, as.numeric(days_employed)),
    years_employed = round(abs(days_employed) / 365, 1),
    credit_income_ratio = amt_credit / amt_income_total,
    annuity_income_ratio = amt_annuity / amt_income_total
  )
```
These transformations do not change the underlying information, but they make the variables easier to interpret in charts and summary statistics. Ratio-based features are also useful because they may capture financial burden better than raw amounts.

## 7. Numeric Feature Exploration
### 7.1 Age Distribution

```{r}
#| label: age-summary
#| echo: true

eda_df |>
  summarise(
    min_age = min(years_birth, na.rm = TRUE),
    median_age = median(years_birth, na.rm = TRUE),
    mean_age = mean(years_birth, na.rm = TRUE),
    max_age = max(years_birth, na.rm = TRUE)
  )
```

```{r}
#| label: age-plot
#| echo: true

ggplot(eda_df, aes(x = years_birth)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Applicant Age Distribution",
    x = "Age in Years",
    y = "Count"
  )
```
The applicant population covers a broad age range. Converting age from negative day counts into years makes the distribution far more interpretable and more useful for business communication.

### 7.2 Income Distribution

```{r}
#| label: income-summary
#| echo: true

eda_df |>
  summarise(
    min_income = min(amt_income_total, na.rm = TRUE),
    median_income = median(amt_income_total, na.rm = TRUE),
    mean_income = mean(amt_income_total, na.rm = TRUE),
    max_income = max(amt_income_total, na.rm = TRUE)
  )
```

```{r}
#| label: income-plot
#| echo: true

ggplot(eda_df, aes(x = amt_income_total)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Income Distribution",
    x = "Total Income",
    y = "Count"
  )
```
Income appears to be highly skewed, which is common in financial data. This suggests that later modeling may benefit from transformations or robust handling of extreme values.

### 7.3 Credit Amount Distribution

```{r}
#| label: credit-summary
#| echo: true

eda_df |>
  summarise(
    min_credit = min(amt_credit, na.rm = TRUE),
    median_credit = median(amt_credit, na.rm = TRUE),
    mean_credit = mean(amt_credit, na.rm = TRUE),
    max_credit = max(amt_credit, na.rm = TRUE)
  )
```

```{r}
#| label: credit-plot
#| echo: true

ggplot(eda_df, aes(x = amt_credit)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Credit Amount Distribution",
    x = "Credit Amount",
    y = "Count"
  )
```
Credit amounts also vary widely across applicants. This suggests that borrower risk may depend not only on the size of the requested credit but also on how that credit compares with income and repayment capacity.

## 8. Categorical Feature Exploration
### 8.1 Contract Type

```{r}
#| label: contract-type
#| echo: true

eda_df |>
  count(name_contract_type) |>
  mutate(percent = n / sum(n)) |>
  arrange(desc(n))
```

```{r}
#| label: contract-type-plot
#| echo: true

eda_df |>
  count(name_contract_type) |>
  ggplot(aes(x = reorder(name_contract_type, n), y = n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Applicants by Contract Type",
    x = "Contract Type",
    y = "Count"
  )
```
This distribution shows the relative frequency of different contract types. It helps identify dominant product categories and provides useful context for understanding the borrower population.

### 8.2 Gender

```{r}
#| label: gender-summary
#| echo: true

eda_df |>
  count(code_gender) |>
  mutate(percent = n / sum(n)) |>
  arrange(desc(n))
```

```{r}
#| label: gender-plot
#| echo: true

eda_df |>
  count(code_gender) |>
  ggplot(aes(x = reorder(code_gender, n), y = n)) +
  geom_col() +
  labs(
    title = "Applicants by Gender",
    x = "Gender",
    y = "Count"
  )
```
Categorical variables such as gender can provide useful descriptive context, although they should be interpreted carefully in any predictive or fairness-related discussion.

## 9. Target Relationship Exploration
### 9.1 Default Rate by Contract Type

```{r}
#| label: default-by-contract
#| echo: true

eda_df |>
  group_by(name_contract_type) |>
  summarise(default_rate = mean(target, na.rm = TRUE), n = n()) |>
  arrange(desc(default_rate))
```

```{r}
#| label: default-by-contract-plot
#| echo: true

eda_df |>
  group_by(name_contract_type) |>
  summarise(default_rate = mean(target, na.rm = TRUE)) |>
  ggplot(aes(x = reorder(name_contract_type, default_rate), y = default_rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title = "Default Rate by Contract Type",
    x = "Contract Type",
    y = "Default Rate"
  )
```
Comparing default rates across groups helps reveal whether some borrower segments may be riskier than others. These patterns can inform both business interpretation and feature engineering.

### 9.2 Default Rate by Income Type

```{r}
#| label: default-by-income-type
#| echo: true

eda_df |>
  group_by(name_income_type) |>
  summarise(default_rate = mean(target, na.rm = TRUE), n = n()) |>
  arrange(desc(default_rate))
```

```{r}
#| label: default-by-income-type-plot
#| echo: true

eda_df |>
  group_by(name_income_type) |>
  summarise(default_rate = mean(target, na.rm = TRUE)) |>
  ggplot(aes(x = reorder(name_income_type, default_rate), y = default_rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title = "Default Rate by Income Type",
    x = "Income Type",
    y = "Default Rate"
  )
```
Borrowers with different income sources may have different repayment patterns. This makes income type a potentially informative feature for downstream risk modeling.

## 10. Feature Engineering Preview

```{r}
#| label: ratio-summary
#| echo: true

eda_df |>
  select(credit_income_ratio, annuity_income_ratio) |>
  summary()
```

```{r}
#| label: ratio-plot
#| echo: true

ggplot(eda_df, aes(x = credit_income_ratio)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Credit-to-Income Ratio Distribution",
    x = "Credit-to-Income Ratio",
    y = "Count"
  )
```

```{r}
#| label: ratio-by-target
#| echo: true

eda_df |>
  group_by(target) |>
  summarise(
    avg_credit_income_ratio = mean(credit_income_ratio, na.rm = TRUE),
    avg_annuity_income_ratio = mean(annuity_income_ratio, na.rm = TRUE)
  )
```
Ratio-based features may be more informative than raw financial values because they better capture repayment burden relative to available income. This is one reason feature engineering is such an important part of the credit risk workflow.

## 11. Key Business Insights

- The strong class imbalance indicates that naive accuracy metrics may be misleading, and alternative evaluation strategies are needed.
- High levels of missing data suggest that data preprocessing decisions will significantly impact model performance.
- Financial burden indicators (such as credit-to-income ratio) are likely more meaningful than raw financial values.
- Different borrower segments (e.g., income type, contract type) show varying default behavior, which can be leveraged for segmentation-based modeling.
- Early insights from EDA directly inform feature engineering and model design decisions.

## 12. Summary

This exploratory data analysis provided an initial understanding of the Home Credit borrower dataset and highlighted several important challenges, including class imbalance, missingness, and skewed financial variables. It also identified useful directions for feature engineering and business interpretation. These findings informed the subsequent data preparation and modeling stages of the project.
